[Artifact Presentation] Damegender: \\ Writing and Comparing Gender Detection Tools

### 2020-05-15 paper2sattose.tex
Ok, so I have read the paper and see what is to be done.

\title{Damegender: \\ Writing and Comparing Gender Detection Tools}

@@

\grex{Here we should have a paragraph introducing the problem}

There are different ways to detect gender from a person name and
perhaps a surname: census, Wikipedia, self-references in trust
websites, searches in Google Images.
Another way to do it is by using one of the existing Application Programming Interfaces (APIs) that exist.
This paper is about the latter, about their possibilities and limitations.
Therefore, (i) we evaluate the quality and price of different commercial solutions;
(ii) we discuss about solutions using free licenses;
(iii) we investigate what happens with those names without census, for example, nicknames or diminutives; and
(iv) we elaborate on how massive gender detection from Internet, for example, mailing
  lists or software repositories, can be done.

As a result, we contribute with: 
(i) an evaluation of the quality of different solutions applying metrics
  suggested by Santamaría and Mihaljević~\cite{10.7717/peerj-cs.156};
(ii) a tool guessing gender from a name giving support to Spanish and
  English from the open data census provides by the states built to understand 
  current technologies in detail;
(iii) a machine learning solution to strings not found in the
  census dataset to approach the problem with nicknames and diminutives; and 
(iv) a proof-of-concept applying Perceval~\cite{duenas2018perceval} to detect
  gender in mailing lists and software repositories.

@@

~\cite{10.7717/peerj-cs.156} are proposing a good metrics set to
classify these commercial Application Programming Interfaces
(features, measuring errors and successes, ...). The features observed
are: Database size (January 2018), Regular data updates, Handles
unstructured full name strings, Handles surnames, Handles non-Latin
alphabets, Implicit geo-localization, Assignment type, Free
parameters, Open source, Application Programming Interface, Monthly
free requests, Monthly subscription cost (100,000 requests/month)).

It's possible different ways to express probability about the
succesful (confidence, scale, accuracy, precision, recall, ...),
confusion matrix to understand where it's successes or fails and
differente errors measures (error coded, error coded without not
applicable values, error gender bias, not applicable coded).

@@

We can divide the next options choosing a dataset: (1) a census
published with a free license (open census way), (2) a dataset done by
scientist with a paper in a magazine (scientific way), (3) a dataset
released with a free license in a free software package (free software
way), (4) a dataset retrieved from commercial Application Programming
Interfaces (commercial api way).

@@

From the user final point of view, the value of using Open Data is
give a good explanation when we are asking about the gender from a
name (number of males and females using a specific name in a country)
versus a probability created by the way explained in
~\cite{10.7717/peerj-cs.156} or similar.

@@

A third approach is using a dataset from a popular free software
solution. For instance, Natural Language Tool Kit is providing 8000
labeled english names. The classification is male or female. The
problem again is about don't retrieve data with the social science
quality of National Statistics Institutes. Another example is Gender
Guesser a good dataset for international names with different
categories to define the probability. This approach is similar to use
a dataset released with a paper in a journal, the advantage is to
understand and add new names with a solid criteria accepted by the
scientific community.

We are using the census approach as base of truth to distinguish if a
name is male or female in a geographical area. Generally, a name has a
strong weight to determine if it's a male or a female on this way, for
instance, David is registered 365196 times as male and 0 times as
female in Spain National Institute of Statistics.

@@

The last approach is based on to trust on commercial solutions, such
as we trust on search engines to make searches in Internet (black
box).  In Damegender we can download json files from main commercial
Application Programming Interfaces (API) solutions (genderapi,
genderize, namsor, nameapi, ...). One user can build proprietary
datasets on this way using an average weighted by the precision or
accuracy of each Application Programming Interface measured with
Damegender with an open dataset as base of truth.

Some Open Datasets, such Spain (INE.es), or United States of America
is providing support to surnames and the relationship with the
ethnicity. For example, in United States of America is giving a
probability with the race and in Spain INE.es is giving the number of
people with a surname with a nationality different to the spanish
nationality. The propietary solutions is allowing infer the gender
with the surname as another parameter, due to names such as Andrea
changes the gender depending the nationality: in Italia would be male,
but in Spain would be female.

@@

In some studies, for example, about Twitter or Github, some people can
choose between different ways to detect gender (not only names). So,
we can find gender detection tools from faces in images
(~\cite{ranjan2017hyperface}), from hand written
(~\cite{liwicki2011automatic}), or from speeches
(~\cite{koppel2002automatically}).

\subsection{Massive Gender Detection}

There are good studies measuring gender in Internet. Some studies are
about gender gap in general (~\cite{robles2014floss},
~\cite{holman2018gender}, ~\cite{dollar1999gender}), Twitter
(~\cite{burger2011discriminating}, ~\cite{mislove2011understanding})
Stackoverflow (~\cite{vasilescu2012gender}), Wikipedia
(~\cite{antin2011gender}, ~\cite{hill2013wikipedia}), Github
(~\cite{vasilescu2015gender}) ...

@@

~\cite{10.7717/peerj-cs.156} explains different ways to determine
gender from a name by humans and it gives 7000 names applying these
methods. In this dataset the gender is classified as male, female or
unknown. We have used this dataset, but only male and female to these
experiments. We are showing the results in the next table:

@@

In ~\ref{table:DifferentAccuraciesMeasures} Genderapi and Genderize
are obtaining the best results, although all solutions is reaching
results better than 0.8 except Gender Guesser.

@@

With Damegender has been done a comparison about confusion matrix
tables depending the API (see ~\ref{table:ConfusionMatrixTables}). If
we compare these results with the results obtained in
~\cite{10.7717/peerj-cs.156}, we can understand that the results are
similar.

Genderapi has similar results, but it is being improved the undefined
results. In Genderguesser is we are obtaining different results and it
is strange, because the software has not modified from some years
ago. In Genderize we are obtaining the same results. In Nameapi the
guessed results is changing from male to female with more errors. In
Namsor the results is so similar. Damegender is not guessing undefined
because we predict with machine learning if the string is not in the
database.

@@

In the table it is possible to observe a high index of errors in
Nameapi and Namsor and a low index of errors in GenderApi
and Damegender.

@@

We have developed a script infofeatures.py with our datasets to
analyze data about features. The datasets used in this experiment was
retrieved from official datasets from national statistical
institutions in Spain, Uruguay, United Kingdom, USA. The features used
are: first letter, last letter, a, b, c, d, e, f, g, h, i, j, k, l, m,
n, o, p, q, r, s, t, u, v, w, x, y, z, vocals, consonants, first
letter, first letter vocal, last letter vocal, last letter consonant,
last letter a. The choosing of features was verified with Principal
Component Analysis.

Take a look to the most informative features results with the
different datasets:

@@

The countries where the main language is spanish (Uruguay + Spain) and
english (USA + United Kingdom + Australia) where is having very
similar variation with the features chosen between males and females
with these datasets (remember that this datasets are being extracted
from official statistics provided by the states). In Canada, a country
french centric has different rules with this features.

The letter a is varying 0.2 from males to females in (USA and Uruguay)
and 0.1 from males to females (United Kingdom, Australia and
Spain). The last letter a is varying 0.5 from males to females in
(Australia, Spain) around 0.4 in (USA, United Kingdom) and 0.2 in
Uruguay. The last letter o from females to males is varying 0.2 in
(Spain, Australia) and is equal in (Uruguay, USA, United Kingdom). For
the last letter consonant all countries is giving the result that is
for males, with results from 0.2 to 0.5: Uruguay and USA (0.5), United
Kingdom (0.4), Australia and Spain (0.3). So last letter vocal is
reverse tha last letter consonant. First letter consonant or first
letter vocal is a non significative feature due to so similar results
in english and spanish.

The success with the different algorithms is showed in the next table:

@@

The results in ~\ref{table:MLAccuracies} shows that using algorithms
as Support Vector Machines or Random Forest against a scientific
dataset created by independent researchers where it is possible to
reach results similar to another commercial solutions about gender
detection tools. Our classifier is binary (only male and female).

We were doing this experiment with NLTK and INE datasets with
accuracies reaching accuracies until 0.745. So it makes sense expect
better results in random datasets augmenting languages and
countries. Due to our solution is not providing arabic or chinesse
alphabets, yet.

So, it's possible infer that Damegender provides a good solution for
nicknames, diminutives, or similar.

@@

We have chosen Python free software tools with a good scientific
impact. Natural Language Toolkit for Natural Language Processing
~\cite{loper2002nltk}. Scikit for Machine Learning
~\cite{pedregosa2011scikit}. Numpy for Numerical Computation
~\cite{van2011numpy}. Matplotlib to visualize results
~\cite{hunter2007matplotlib}. And Perceval~\cite{duenas2018perceval}
to retrieve information in mailing lists and repositories.

>>>>

\title{[Artifact Presentation] Damegender: \\ Writing and Comparing Gender Detection Tools}

@@

In recent times, many research investigations have been made on gender
diversity in the IT domain.
Examples of these efforts range from participation in Twitter~\cite{burger2011discriminating,mislove2011understanding},
in Wikipedia ~\cite{antin2011gender,hill2013wikipedia}, in science~\cite{holman2018gender,dollar1999gender}, and more specifically in the software domain in StackOverflow~\cite{vasilescu2012gender}, GitHub~\cite{vasilescu2015gender} and in Free/Libre/Open Source Software development~\cite{robles2014floss}.
The interest on gender diversity is become more and more relevant, and so does the identification methods that allow to perform comprehensive studies on gender representation in different domains, given the large amounts of data available, in particular from collaborative environments.

There are different ways to detect gender from a person name and perhaps a surname.
A first, more rudimentary, is based on data extracted from the census, Wikipedia, self-references in trust websites, searches in Google Images, among others.
Another way to do it is by using one of the existing Application Programming Interfaces (APIs).
This paper is about the latter, about their possibilities and limitations.
Therefore, (i) we evaluate the quality and price of different commercial solutions;
(ii) we discuss about solutions using free licenses;
(iii) we investigate what happens with those names without census, for example, nicknames or diminutives; and
(iv) we elaborate on how massive gender detection from Internet, for example, mailing
  lists or software repositories, can be done.

As a result, we contribute with: 
(i) an evaluation of the quality of different solutions applying well-known metrics;
(ii) a tool, called \texttt{damegender}, guessing gender from a name giving support to Spanish and
  English from the open data census provides by the states built to understand 
  current technologies in detail;
(iii) a machine learning solution to strings not found in the
  census dataset to approach the problem with nicknames and diminutives; and 
(iv) a proof-of-concept of \texttt{damegender} to detect
  gender in mailing lists and software repositories.

@@

Santamara and Mihaljevi~\cite{10.7717/peerj-cs.156} propose a set of metrics to classify gender tools (features, measuring errors and successes, among others).
The features observed are: (i) database size (as of January 2018), (ii) if there are regular data updates, (iii) if they handle
unstructured full name strings, (iv) if they handles surnames, (v) if they handles non-Latin alphabets, (vi) if implicit geolocalization is available, (vii) the type of assignment, (viii) if free parameters are possible, (ix) if the tool is open source, (x) if they offer an API, (xi) the amount of monthly free requests, and (xii) the monthly subscription cost (calculated for 100,000 requests/month)).

There are different ways to express the probability of a successful identification (confidence, scale, accuracy, precision, recall, ...), confusion matrix to understand where it is successes or fails and different errors measures (error coded, error coded without not applicable values, error gender bias, not applicable coded).

@@

We can divide the next options choosing a dataset: (1) a census published with a free license (open census way), (2) a dataset done by scientist with a paper in a magazine (scientific way), (3) a dataset released with a free license in a free software package (free software way), (4) a dataset retrieved from commercial APIs (commercial API way).

@@

From the user final point of view, the value of using Open Data is give a good explanation when we are asking about the gender from a name (number of males and females using a specific name in a country) versus a probability created by the way explained in~\cite{10.7717/peerj-cs.156} or similar.

@@

A third approach is using a dataset from a popular free software solution. For instance, Natural Language Tool Kit is providing 8000 labeled English names. 
The classification is male or female. The problem again is about do not retrieve data with the social science quality of National Statistics Institutes. 
Another example is Gender Guesser a good dataset for international names with different categories to define the probability. 
This approach is similar to use a dataset released with a paper in a journal, the advantage is to understand and add new names with a solid criteria accepted by the scientific community.

We are using the census approach as base of truth to distinguish if a name is male or female in a geographical area. Generally, a name has a strong weight to determine if it is a male or a female on this way.
For instance, David is registered 365196 times as male and 0 times as female in Spain National Institute of Statistics.

@@

The last approach is based on to trust on commercial solutions, such as we trust on search engines to make searches in Internet (black box).
In Damegender we can download JSON files from main commercial Application Programming Interfaces (API) solutions (genderapi,
genderize, namsor, nameapi, ...).
One user can build proprietary datasets on this way using an average weighted by the precision or accuracy of each Application Programming Interface measured with Damegender with an open dataset as base of truth.

Some Open Datasets, such Spain (INE.es), or United States of America is providing support to surnames and the relationship with the ethnicity. 
For example, in United States of America is giving a probability with the race and in Spain INE.es is giving the number of
people with a surname with a nationality different to the Spanish nationality.
The proprietary solutions is allowing infer the gender with the surname as another parameter, due to names such as Andrea
changes the gender depending the nationality: in Italy would be male, but in Germany it would be female.

@@

In some studies, for example, about Twitter or GitHub, some people can choose between different ways to detect gender (not only names). 
So, we can find gender detection tools from faces in images~\cite{ranjan2017hyperface}, from hand written annotations~\cite{liwicki2011automatic}, or from speeches~\cite{koppel2002automatically}.

@@

~\cite{10.7717/peerj-cs.156} explains different ways to determine gender from a name by humans and it gives 7000 names applying these methods. 
In this dataset the gender is classified as male, female or unknown. 
We have used this dataset, but only male and female to these experiments.
We are showing the results in the Table~\ref{table:DifferentAccuraciesMeasures}.

@@

In Table~\ref{table:DifferentAccuraciesMeasures}, Genderapi and Genderize are obtaining the best results, although all solutions is reaching results better than 0.8 except Gender Guesser.

@@

With Damegender has been done a comparison about confusion matrix tables depending the API (see Table~\ref{table:ConfusionMatrixTables}).
If we compare these results with the results obtained in~\cite{10.7717/peerj-cs.156}, we can understand that the results are
similar.

Genderapi has similar results, but it is being improved the undefined results.
In Genderguesser is we are obtaining different results and it is strange, because the software has not modified from some years
ago. In Genderize we are obtaining the same results. 
In Nameapi the guessed results is changing from male to female with more errors. 
In Namsor the results is similar. 
Damegender is not guessing undefined because we predict with machine learning if the string is not in the database.

@@

In Table~\ref{table:ApisAndErrors} we can observe a high index of errors in Nameapi and Namsor and a low index of errors in GenderApi and Damegender.

@@

We have developed a script infofeatures.py with our datasets to analyze data about features.
The datasets used in this experiment was retrieved from official datasets from national statistical
institutions in Spain, Uruguay, United Kingdom, USA. The features used are: first letter, last letter, [a-z], vocals, consonants, first letter, first letter vocal, last letter vocal, last letter consonant, last letter a. 
The choosing of features was verified with Principal Component Analysis.
The most relevant results for the different datasets are offered in Table~\ref{table:InfoFeatures}.

@@

The countries where the main language is Spanish (Uruguay + Spain) and English (USA + United Kingdom + Australia) where is having very similar variation with the features chosen between males and females with these datasets (remember that these datasets are being extracted from official statistics provided by the states). 
In Canada, a country French centric has different rules with this features.

The letter a is varying 0.2 from males to females in (USA and Uruguay) and 0.1 from males to females (United Kingdom, Australia and Spain).
The last letter a is varying 0.5 from males to females in (Australia, Spain) around 0.4 in (USA, United Kingdom) and 0.2 in
Uruguay.
The last letter o from females to males is varying 0.2 in (Spain, Australia) and is equal in (Uruguay, USA, United Kingdom).
For the last letter consonant all countries is giving the result that is for males, with results from 0.2 to 0.5: Uruguay and USA (0.5), United Kingdom (0.4), Australia and Spain (0.3). 
So last letter vocal is reverse than last letter consonant. 
First letter consonant or first letter vocal is a non significant feature due to so similar results in English and Spanish.
The relevant metrics with the different algorithms are shown in Table~\ref{table:MLAccuracies}.

@@

The results in Table~\ref{table:MLAccuracies} show that using algorithms as Support Vector Machines or Random Forest against a scientific dataset created by independent researchers where it is possible to reach results similar to another commercial solutions about gender detection tools.
Our classifier is binary (only male and female).

We were doing this experiment with NLTK and INE datasets with accuracies reaching accuracies until 0.745.
So it makes sense expect better results in random datasets augmenting languages and countries.
However, our solution is not providing Arabic or Chinese alphabets, yet.

So, it is possible to infer that Damegender provides a good solution for nicknames, diminutives, or similar.

@@

We have chosen Python free software tools with a good scientific impact.
Natural Language Toolkit for Natural Language Processing ~\cite{loper2002nltk}.
Scikit for Machine Learning~\cite{pedregosa2011scikit}. 
Numpy for Numerical Computation~\cite{van2011numpy}. 
Matplotlib to visualize results~\cite{hunter2007matplotlib}. 
And Perceval~\cite{duenas2018perceval} to retrieve information in mailing lists and repositories.
